Bayesian Rate Estimation

### 2012-08-13 many.tex
Removed red from Sec. II.

\author{Jon Gair}
\email{jrg23@cam.ac.uk}

@@

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process \ilya{[pardon my ignorance, but
    I thought an "inhomogeneous" Poisson process meant that the rate
    changed as a function of time -- and I don't think that's what we
    are assuming...]} with \ilya{respective} differential rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and 
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent. 
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(\ilya{n} = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligible
  \begin{equation}
    P(\ilya{n} = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and 
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f\ilya{\equiv
  F(\infty,\theta)}$ and $R_b\ilya{\equiv B(\infty,\theta)}$ are the
total number of foreground and background events expected and
$\hat{f}(x, \theta')$ and $\hat{b}(x, \theta')$ are the likelihood of
obtaining an event with ranking statistic $x$ under the foreground and
background distributions.  In what follows, we will drop the prime,
using $\theta$ to denote all parameters of the rate distributions
except $R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 \ilya{(background)} or 1 \ilya{(foreground)}.  These
``state'' flags are parameters in our model, along with $R_f$, $R_b$,
and $\theta$.  We can marginalize over our uncertainty in the state of
any given event by summing posteriors over $f_i = \mathset{0,1}$.

Bayes' theorem relates the posterior probability of the state flags,
rates, and shape parameters, $p\left(\mathset{f_i}, R_f, R_b, \theta |
  d \right)$, the likelihood of the data, $p\left( d | \mathset{f_i},
  R_f, R_b, \theta\right)$, and the prior probability of state flags,
rates and shape parameters before any data are obtained, $p\left(
  \mathset{f_i}, R_f, R_b, \theta\right)$:
\begin{equation}
  \label{eq:Bayes}
  p\left(\mathset{f_i}, R_f, R_b, \theta | d \right)  \ilya{=}
  \frac{p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)}{\ilya{p(d)}}. 
\end{equation}
The \ilya{normalization constant}, called the evidence, $p(d)$, is
independent of the state flags, rates, and shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is 
\begin{equation}
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)  = \left[
    \prod_{\mathset{i | f_i = 1}} \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} \hat{b}\left( x_i, \theta\right)
  \right].  
\end{equation}
The prior distribution can be factorized as
\begin{equation}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)  = p\left(
    \mathset{f_i} | R_f, R_b, \theta\right)p(R_f, R_b, \theta).
\end{equation}
The first term is not actually a ``prior'' in the usual sense; the
conditional probability of the flags $\mathset{f_i}$ on the rates is given
by
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{f_i} | R_f, R_b, \theta\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N_f! N_b!} \exp\left[ - \left(R_f + R_b\right) \right]
  N_f! N_b! = R_f^{N_f}R_b^{N_b} \exp\left[ - \left(R_f + R_b\right) \right],
\end{equation}
where $N_f$ and $N_b$ are the number of foreground and background
events.  This expression follows from Property \ref{prop:Poisson} of
inhomogeneous Poisson processes and the $N_f! N_b!$ different ways of
assigning foreground and background flags to events for fixed $N_f$,
$N_b$.

\ilya{[I agree with the final result, but I find this way of deriving
    it to be confusing.  First of all, the number of ways to assign
    flags to events for fixed $N_f$ and $N_b$ is $N!/(N_f! N_b!)$, not
    $N_f! N_b!$ as written.  Secondly, you are computing the
    probability of having a particular ordered list of flags, so
    there's no reason to consider different permutations.  Thirdly,
    you are not properly including the fact that the length of the
    flags vector must be exactly $N$, the total number of data points.
    (Well, you have the right result, so you are obviously doing these
    things properly, but the explanation just doesn't seem clear to
    me. :) ) I would make that $N$ explicit, by writing Eq.~(12) as

$$
  p\left(\mathset{f_i}, N, R_f, R_b, \theta\right)  = 
  p\left( \mathset{f_i} | N, R_f, R_b\right) p\left(N |R_f, R_b\right)  
 p\left(R_f, R_b, \theta\right).
$$

Then 
$$ p\left( \mathset{f_i} | N, R_f, R_b\right) = 
\prod_{\mathset{i|f_i=1}} \left(\frac{R_f}{R_f+R_b}\right) 
\prod_{\mathset{i|f_i=0}} \left(\frac{R_b}{R_f+R_b}\right) =
\left(\frac{R_f}{R_f+R_b}\right)^{N_f} \left(\frac{R_b}{R_f+R_b}\right)^{N_b}. 
$$ Note that there are no permutations involved here -- once we assume
that there are precisely $N=N_f+N_b$ events, the probability for the
i'th event to have flag $f_i$ is either $R_f/(R_f+R_b)$ or
$R_b/(R_f+R_b)$ depending on whether $f_i$ is 1 or 0.  Meanwhile,
$$
p\left(N |R_f, R_b\right) = \frac{\left(R_f+R_b\right)^N}{N!} e^{-(R_f+R_b)},
$$ since the distribution of total event number is a Poisson process
with rate $R_f+R_b$.  Multiplying these, we get the same result as in
(13), but in a way that even I can understand. :) ]}

The second term in Eq.~\eqref{eq:combined-flag-rate-prior} is a
traditional prior.  Because the rate parameters enter the posterior in
the same form as Poisson rates, we choose here the Poisson Jeffreys
prior on rates \citep{Jeffreys1946}, independent of the shape
parameters
\begin{equation}
  p\left( R_f, R_b, \theta\right) \ilya{\propto} \frac{1}{\sqrt{R_f R_b}} p(\theta),
\end{equation}
but of course other choices are possible.  This choice has the
advantage that the prior is normalizable as $R_f, R_b \to 0$, and the
exponentials in Eq.~\eqref{eq:flag-conditional-prior} regularize the
posterior as $R_f, R_b \to \infty$.

@@

Eq.~\eqref{eq:posterior} is unchanged if the ranking statistic is
multi-dimensional; in this case, the rates are 
\begin{equation}
  R_f = \int d^\ilya{k} \ilya{\vec{x}} \, f(x, \theta)
\end{equation}
and
\begin{equation}
  R_b = \int d^\ilya{k} \ilya{\vec{x}} \, b(x, \theta),  
\end{equation}
where $f$ and $b$ are rate densities on the \ilya{$k$}-dimensional
space of ranking statistics.  We give an example of fitting for
multi-dimensional rate densities in \S~\ref{sec:star-cluster}.

@@

To analyize our synthetic data set, we analytically marginalized over
the state \ilya{[this is the only time we call them "state" flags...]}
flags, using the likelihood in Eq.~\eqref{eq:rate-shape-posterior}.
\ilya{[I think it would be kind to the reader to be more explicit
    about the connection with the formalism in Section II.  What are
    the "events"?  What is the ranking statistic $x$?  What are the
    flags (which aren't in $\mathbb{R}$?)?]}  We did this to take
advantage of the \textit{emcee} sampler of \citet{ForemanMackey2012},
which requires all parameters to be in $\mathbb{R}$.  We applied a
prior on the shape parameters that is flat in $\vec{x}_0$ and
$\vec{\gamma}$, and the Jeffreys prior on $r_0$,
\begin{equation}
  p\left( r_0 \right) = \frac{\sqrt{R_f}}{r_0}.
\end{equation}
\ilya{[Why is this the Jeffreys prior (BTW, it's Jeffreys, or perhaps
    Jeffreys', but definitely not Jeffrey's ;) )?  It's not obvious
    why the prior on $r_0$ should depend on $R_f$?]}  (Note that this
factor of $\sqrt{R_f}$ cancels with the Jeffreys prior on the rate,
$1/\sqrt{R_f}$; we have verified that the priors on these parameters
are irrelevant to our results, as would be expected from the
measurement of $\sim 1000$ foreground stars.)

>>>>

\author{Jonathan R. Gair}
\email{jrg23@cam.ac.uk}

@@

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process with respective differential rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and 
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent. 
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(n = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligible
  \begin{equation}
    P(n = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and 
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f \equiv
  F(\infty,\theta)$ and $R_b \equiv B(\infty,\theta)$ are the
total number of foreground and background events expected and
$\hat{f}(x, \theta')$ and $\hat{b}(x, \theta')$ are the likelihood of
obtaining an event with ranking statistic $x$ under the foreground and
background distributions.  In what follows, we will drop the prime,
using $\theta$ to denote all parameters of the rate distributions
except $R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 (background) or 1 (foreground).  These
``state'' flags are parameters in our model, along with $R_f$, $R_b$,
and $\theta$.  We can marginalize over our uncertainty in the state of
any given event by summing posteriors over $f_i = \mathset{0,1}$.

Bayes' theorem relates the posterior probability of the state flags,
rates, and shape parameters, $p\left(\mathset{f_i}, R_f, R_b, \theta |
  d \right)$, the likelihood of the data, $p\left( d | \mathset{f_i},
  R_f, R_b, \theta\right)$, and the prior probability of state flags,
rates and shape parameters before any data are obtained, $p\left(
  \mathset{f_i}, R_f, R_b, \theta\right)$:
\begin{equation}
  \label{eq:Bayes}
  p\left(\mathset{f_i}, R_f, R_b, \theta | d \right)  =
  \frac{p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)}{p(d)}. 
\end{equation}
The normalization constant, called the evidence, $p(d)$, is
independent of the state flags, rates, and shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is 
\begin{equation}
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)  = \left[
    \prod_{\mathset{i | f_i = 1}} \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} \hat{b}\left( x_i, \theta\right)
  \right].  
\end{equation}

%The prior distribution can be factorized as
%\begin{equation}
%  \label{eq:combined-flag-rate-prior}
%  p\left(\mathset{f_i}, R_f, R_b, \theta\right)  = p\left(
%   \mathset{f_i} | R_f, R_b, \theta\right)p(R_f, R_b, \theta).
%\end{equation}
%The first term is not actually a ``prior'' in the usual sense; the
%conditional probability of the flags $\mathset{f_i}$ on the rates is given
%by
%\begin{equation}
%  \label{eq:flag-conditional-prior}
%  p\left(\mathset{f_i} | R_f, R_b, \theta\right) = \frac{R_f^{N_f}
%    R_b^{N_b}}{N_f! N_b!} \exp\left[ - \left(R_f + R_b\right) \right]
%  N_f! N_b! = R_f^{N_f}R_b^{N_b} \exp\left[ - \left(R_f + R_b\right) \right],
%\end{equation}
%where $N_f$ and $N_b$ are the number of foreground and background
%events.  This expression follows from Property \ref{prop:Poisson} of
%inhomogeneous Poisson processes and the $N_f! N_b!$ different ways of
%assigning foreground and background flags to events for fixed $N_f$,
%$N_b$.  
%
%\ilya{[I agree with the final result, but I find this way of deriving
%    it to be confusing.  First of all, the number of ways to assign
%    flags to events for fixed $N_f$ and $N_b$ is $N!/(N_f! N_b!)$, not
%    $N_f! N_b!$ as written.  Secondly, you are computing the
%    probability of having a particular ordered list of flags, so
%    there's no reason to consider different permutations.  Thirdly,
%    you are not properly including the fact that the length of the
%    flags vector must be exactly $N$, the total number of data points.
%    (Well, you have the right result, so you are obviously doing these
%    things properly, but the explanation just doesn't seem clear to
%    me. :) ) I would make that $N$ explicit, by writing Eq.~(12) as

The prior distribution can be factorized as
\begin{equation}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{f_i}, N, R_f, R_b, \theta\right)  = 
  p\left( \mathset{f_i} | N, R_f, R_b\right) p\left(N |R_f, R_b\right)  
 p\left(R_f, R_b, \theta\right).
 \end{equation}

The {\it a priori} probability that the $i$'th state flag is $f_i=1$ is given by $R_f/(R_f+R_b)$, while the probability that it is zero is $R_b/(R_f+R_b)$.  Then
\begin{equation}
p\left( \mathset{f_i} | N, R_f, R_b\right) = 
\prod_{\mathset{i|f_i=1}} \left(\frac{R_f}{R_f+R_b}\right) 
\prod_{\mathset{i|f_i=0}} \left(\frac{R_b}{R_f+R_b}\right) =
\left(\frac{R_f}{R_f+R_b}\right)^{N_f} \left(\frac{R_b}{R_f+R_b}\right)^{N_b},
\end{equation}
where $N_f$ and $N_b$ are the numbers of foreground and background flags, $N_f+N_b=N$.
Meanwhile,
\begin{equation}
p\left(N |R_f, R_b\right) = \frac{\left(R_f+R_b\right)^N}{N!} e^{-(R_f+R_b)},
\end{equation}
since the distribution of total event number is a Poisson process
with rate $R_f+R_b$.  Combining these yields the conditional probability of the flags on the rates:
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{f_i} | R_f, R_b\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N_f! N_b!} \exp\left[ - \left(R_f + R_b\right) \right]
  N_f! N_b! = R_f^{N_f}R_b^{N_b} \exp\left[ - \left(R_f + R_b\right) \right].
\end{equation}

The second term in Eq.~\eqref{eq:combined-flag-rate-prior} is a
traditional prior.  Because the rate parameters enter the posterior in
the same form as Poisson rates, we choose here the Poisson Jeffreys
prior on rates \citep{Jeffreys1946}, independent of the shape
parameters
\begin{equation}
  p\left( R_f, R_b, \theta\right) \propto \frac{1}{\sqrt{R_f R_b}} p(\theta),
\end{equation}
but of course other choices are possible.  This choice has the
advantage that the prior is normalizable as $R_f, R_b \to 0$, and the
exponentials in Eq.~\eqref{eq:flag-conditional-prior} regularize the
posterior as $R_f, R_b \to \infty$.

@@

Eq.~\eqref{eq:posterior} is unchanged if the ranking statistic is
multi-dimensional; in this case, the rates are 
\begin{equation}
  R_f = \int d^k \vec{x} \, f(x, \theta)
\end{equation}
and
\begin{equation}
  R_b = \int d^k \vec{x} \, b(x, \theta),  
\end{equation}
where $f$ and $b$ are rate densities on the $k$-dimensional
space of ranking statistics.  We give an example of fitting for
multi-dimensional rate densities in \S~\ref{sec:star-cluster}.

@@

To analyize our synthetic data set, we analytically marginalized over
the state flags, using the likelihood in Eq.~\eqref{eq:rate-shape-posterior}.
\ilya{[I think it would be kind to the reader to be more explicit
    about the connection with the formalism in Section II.  What are
    the "events"?  What is the ranking statistic $x$?  What are the
    flags (which aren't in $\mathbb{R}$?)?]}  We did this to take
advantage of the \textit{emcee} sampler of \citet{ForemanMackey2012},
which requires all parameters to be in $\mathbb{R}$.  We applied a
prior on the shape parameters that is flat in $\vec{x}_0$ and
$\vec{\gamma}$, and the Jeffreys prior on $r_0$,
\begin{equation}
  p\left( r_0 \right) = \frac{\sqrt{R_f}}{r_0}.
\end{equation}
\ilya{[Why is this the Jeffreys prior (BTW, it's Jeffreys, or perhaps
    Jeffreys', but definitely not Jeffrey's ;) )?  It's not obvious
    why the prior on $r_0$ should depend on $R_f$?]}  (Note that this
factor of $\sqrt{R_f}$ cancels with the Jeffreys prior on the rate,
$1/\sqrt{R_f}$; we have verified that the priors on these parameters
are irrelevant to our results, as would be expected from the
measurement of $\sim 1000$ foreground stars.)
